Do moral communities have a spatial dimension? A spatial exploratory analysis of places of worship and violent crime in the city of Recife, Brazil

### 2020-06-13 Moral-Communities-and-Crime/Response-to-Reviewers-v2.Rmd

>>\textcolor{blue}{2. Several types of community organizations, including schools, clubs, and neighborhood associations, can strengthen the capabilities of communities to exert informal social control.}

>>\textcolor{blue}{3. Of several types of community organizations, churches are among the most overtly moral, since they often have prescriptive codes of behavior.}

>>\textcolor{blue}{4. Why would the location of places of worship matter? Groff (2015) reviews several theoretical perspectives that address the role of informal social control in micro-places: these include "eyes on the street" and human territorial functioning. By affecting informal surveillance, movement governing rules, direct intervention, and the social climate at places, places of worship are hypothesized to contribute to informal social controls.}

>\textcolor{blue}{We believe that this argument is logical and sound, and churches, in fact, have been studied in the past for their potential to suppress crime (Abdullah, 2018; Davignon, 2015; Furr, 2010; Traunmuller, 2011). A more novel aspect is the study of places of worship, as distinct entities from the members of the congregations: as Warner and Konkel (2019) note, this is an aspect that has received less attention in empirical and theoretical research.}

@@

>\textcolor{blue}{The second question is related to the above: if places of worship somehow help to strengthen social control, does this happen at the level of points or for areas? In our view, Groff (2015) already explains why these investigations should matter at the micro-level. But there is also an analytical aspect to this question: what is an appropriate unit of support for the analysis? In this sense, we would note that aggregating point data to areas inevitably involves loss of information. Sometimes this situation is unavoidable (say, statistics are recorded for areas, so locational information is lost from the origin.) But when we have access to detailed geographical information, converting to area data is an unnecessary loss of information. The question that we are asking is actually quite geographical: if there is evidence of a moral community effect, what is its distance decay? This question is best answered by considering the distance from criminal events to places of worship, instead of simple propinquity of criminal events and places of worship in areas.}

>\textcolor{blue}{We have edited the paper to increase its clarity.}

The current framing of the article also does not convincingly situate 'moral communities' within a geographical framework and does not make clear the hypothesized mechanisms that link the presence/absence of a church at a specific location with criminal behavior and, more specifically, the presence/absence of churches of specific religions with violent crimes. The authors should make sure not to conflate the processes linking crime with population characteristics (percent of individuals who are religious in an area), with religion more broadly (hellfire hypothesis), and with the location of a church in geographic space. This theoretical work is necessary to justify why this study is relevant and make clear how it contributes to the existing literature.

>\textcolor{blue}{Thank you for this comment. As pointed out above, the theoretical work was discussed by Groff (2015), among others. Although conceptually the study draws from much earlier research (e.g., Appleyard, 1980; Jacobs, 1961, 1968; Taylor, 1988, 1998), empirical work has lagged, as recently noted by Warner and Konkel (2019). This is where the current paper aims to make a contribution.}

@@

There’s a cultural issue which I find challenging.  I think I know what Catholic and Evangelical churches may be, but this is first encounter with Spiritism. I assume that the first two are differentiated along some transubstantiation/consubstantiation divide, with the Catholics being in the former group. As for Spiritist churches I’m at a loss.  Do these follow the teaching of Allan Kardec; I presume it’s related to the Spiritualist churches found in the Anglophone countries, which go back to Swedenborg and Mesmer, and which seem to be originally motivated by trying to make contact with the dead. If so, then it’s very much a minority sect.  Perhaps the authors might like to expand on the details further for those readers who are not familiar with religious practices, and how they are undertaken in Brazil. Would I recognise the same in Cape Town, to take a city at random. Would this be a typical Spiritist church: https://goo.gl/maps/kwWe2B7didMk8vfDA

>\textcolor{blue}{We have included in the text some additional information of the religions studied in the article.}

Some description of the geography of Recife would also help. Where are residential areas, where are industrial zones, is there a university, and so on.

>\textcolor{blue}{We now include a figure with maps for socio-economic and demographic variables which help to provide additional context.}

### 2020-06-13 Moral-Communities-and-Crime/Moral_Communities_and_Crime.Rmd
v2 Second submission after comments from Edivaldo and Tatiane

The case study is the city of Recife in the state of Pernambuco, in Brazil's Northeast. Recife is a large and important metropolitan area in a historically poor region, and afflicted by high levels of violent crime, having the dubious honor of being one of five state capitals with the highest rates of homicide in the period under study [@Menezes2013spatial]. The empirical strategy is to use spatial analysis to explore the potential geographical relationships between violent criminal events, on the one hand, and places of worship and a selection of commercial establishments that serve as controls, on the other.

@@

The data are organized at the individual level and it is possible to obtain information about the crime event location, day of the week, day of the month, period of the day, as well as gender, age, and race of the victim. The database used in this study comprises the period from July 1, 2008 to June 30, 2010. Some descriptive statistics regarding this dataset are reported in Table 1, where it can be seen that overall, about 91% of the victims of LIVC in the period analyzed were men, while approximately 98% of violent crimes were homicides. In addition, most of the victims were black or brown, and the youth population between the ages of 18 and 30 is the most affected by violent crime. Lastly, it should be noted that about 88% of the criminal events under analysis involved firearms. Figure \ref{fig:plot-crime} shows the spatial distribution of the 1,657 LIVC crimes that occurred in the city of Recife between July, 2008 and June, 2010.

@@

sed_descriptives <- data.frame(variable = c("Median Income", 
                                            "Proportion Unemployed", 
                                            "Proportion Poverty", 
                                            "Proportion Single Mother", 
                                            "Proportion Young Single Mother",
                                            "Population Density"),
                               min = c(min(setores.sf$median_income),
                                       min(setores.sf$prop_unemployed),
                                       min(setores.sf$prop_poverty),
                                       min(setores.sf$prop_single_mother),
                                       min(setores.sf$prop_young_single_mother),
                                       min(setores.sf$pop_density)),
                               second_q = c(quantile(setores.sf$median_income, c(0, 0.25, 0.5, 0.75, 1))[2],
                                            quantile(setores.sf$prop_unemployed, c(0, 0.25, 0.5, 0.75, 1))[2],
                                            quantile(setores.sf$prop_poverty, c(0, 0.25, 0.5, 0.75, 1))[2],
                                            quantile(setores.sf$prop_single_mother, c(0, 0.25, 0.5, 0.75, 1))[2],
                                            quantile(setores.sf$prop_young_single_mother, c(0, 0.25, 0.5, 0.75, 1))[2],
                                            quantile(setores.sf$pop_density, c(0, 0.25, 0.5, 0.75, 1))[2]),
                               median = c(median(setores.sf$median_income),
                                          median(setores.sf$prop_unemployed),
                                          median(setores.sf$prop_poverty),
                                          median(setores.sf$prop_single_mother),
                                          median(setores.sf$prop_young_single_mother),
                                          median(setores.sf$pop_density)),
                               mean = c(mean(setores.sf$median_income),
                                        mean(setores.sf$prop_unemployed),
                                        mean(setores.sf$prop_poverty),
                                        mean(setores.sf$prop_single_mother),
                                        mean(setores.sf$prop_young_single_mother),
                                        mean(setores.sf$pop_density)),
                               third_q = c(quantile(setores.sf$median_income, c(0, 0.25, 0.5, 0.75, 1))[4],
                                           quantile(setores.sf$prop_unemployed, c(0, 0.25, 0.5, 0.75, 1))[4],
                                           quantile(setores.sf$prop_poverty, c(0, 0.25, 0.5, 0.75, 1))[4],
                                           quantile(setores.sf$prop_single_mother, c(0, 0.25, 0.5, 0.75, 1))[4],
                                           quantile(setores.sf$prop_young_single_mother, c(0, 0.25, 0.5, 0.75, 1))[4],
                                           quantile(setores.sf$pop_density, c(0, 0.25, 0.5, 0.75, 1))[4]),
                               max = c(max(max(setores.sf$median_income),
                                           max(setores.sf$prop_unemployed),
                                           max(setores.sf$prop_poverty),
                                           max(setores.sf$prop_single_mother),
                                           max(setores.sf$prop_young_single_mother),
                                           setores.sf$pop_density)),
                               pca_loadings = as.numeric(setores.pca$rotation[c(2, 3, 6, 4, 5, 1), 1]))
```

@@

As seen in the table, Recife is a city with large socio-economic and demographic disparities; for example, the _setor_ with the highest median income has a median income that is `r prettyNum(filter(sed_descriptives, variable == "Median Income")["max"]/filter(sed_descriptives, variable == "Median Income")["min"] * 100, digits = 2, big.mark = ",")`% higher than the median income in the _setor_ with the lowest median income. Whereas there are _setores_ where the proportion of population who are unemployed is zero, there are _setores_ where almost 60% of the population are unemployed. Likewise, there are _setores_ where almost 70% of the population lives in poverty. In addition to these economic indicators, two variables are used to represent family disruption, the proportion of families whose head is a single mother, and the proportion of families whose head is a _young_ single mother, that is, a woman between the ages of 15 and 25. As can be seen, there are _setores_ with approximately 70% of households led by single mothers, and of these, over 21% are led by younger women, indicating a high degree of family disruption. Unlike other places, where higher population density is associated with more poverty, the distribution of this variable in Recife is more mixed: middle income and higher income households favor high density development in many parts of the city. The spatial distribution of these variables can be seen in Figure \ref{fig:plot-sed-variables}.

@@

Once the indicator of Socio-Economic Deprivation has been obtained, as outlined above, it can be used for conditional quadrat analysis. As previously described, the quadrats are irregular areas that are defined based on a covariate, in this case the quintiles of the Socio-Economic Deprivation indicator. Each color in Figure \ref{fig:plot-sed-as-quintiles} is a "container" for counting events. In essence, this entails counting the number of events at each level of the covariate, and then calculating an estimator of the intensity. The global intensity of each type of event (i.e., places of worship and businesses), as well as quadrant counts and intensity at each level of SED can be consulted in Table \ref{tab:quadrat-count-results}.

@@

intensity_all <- data.frame(SED = rep(c("Low", 
                                        "Middle Low", 
                                        "Middle", 
                                        "Middle High", 
                                        "High"), 
                                      times = 7),
                            Type = rep(c("Violent Crime", 
                                         "Catholic",
                                         "Evangelical",
                                         "Spiritist",
                                         "Ice Cream",
                                         "Pharmacies",
                                         "Super Markets"), 
                                       each = 5),
                            Quadrat_Count = c(crime5,
                                              catholic5,
                                              evangelic5,
                                              spiritist5,
                                              ice_cream5,
                                              pharmacies5,
                                              super_markets5),
                            Intensity = c(crime5_intensity,
                                          catholic5_intensity,
                                          evangelic5_intensity,
                                          spiritist5_intensity,
                                          ice_cream5_intensity,
                                          pharmacies5_intensity,
                                          supermarkets5_intensity),
                            Global_Intensity = rep(c(summary_crime$intensity,
                                                     summary_catholic$intensity,
                                                     summary_evangelic$intensity,
                                                     summary_spiritist$intensity,
                                                     summary_ice_cream$intensity,
                                                     summary_pharmacy$intensity,
                                                     summary_supermarket$intensity),
                                                   each = 5)) %>%
  mutate(SED = factor(SED, levels = c("Low", 
                                      "Middle Low", 
                                      "Middle", 
                                      "Middle High", 
                                      "High"),
                      ordered = TRUE),
         Type = factor(Type, levels = c("Violent Crime", 
                                        "Catholic",
                                        "Evangelical",
                                        "Spiritist",
                                        "Ice Cream",
                                        "Pharmacies",
                                        "Super Markets")))
```

```{r table-counts-intensity-summary, echo=FALSE}
kable(intensity_all %>% dplyr::select(Type, Global_Intensity, SED, Quadrat_Count, Intensity),
      "latex",
      booktabs = TRUE,
      digits = 3,
      escape = FALSE,
      align = c("l", "c", "c", "c"),
      caption = "\\label{tab:quadrat-count-results}Conditional quadrat counts results",
      col.names = linebreak(c("Event", 
                              "Global Intensity", 
                              "SED Quintile", 
                              "Quadrat Count",
                              "Intensity"))) %>%
  kable_styling(bootstrap_options = c("striped", "scaled_down")) %>% 
  collapse_rows(1:2, latex_hline = "none")
```

>>>>

The empirical analysis uses geographically detailed data from the city of Recife in the state of Pernambuco, in Brazil's Northeast. Recife is a large and important metropolitan area in a historically poor region, and afflicted by high levels of violent crime, having the dubious honor of being one of five state capitals with the highest rates of homicide in the period under study [@Menezes2013spatial]. The empirical strategy is to use spatial analysis to explore the potential geographical relationships between violent criminal events, on the one hand, and places of worship and a selection of commercial establishments that serve as controls, on the other.

@@

The data are organized at the individual level and it is possible to obtain information about the crime event location, day of the week, day of the month, period of the day, as well as gender, age, and race of the victim. The database used in this study comprises the period from July 1, 2008 to June 30, 2010. Some descriptive statistics regarding this dataset are reported in Table 1, where it can be seen that overall, about 91% of the victims of LIVC in the period analyzed were men. In addition, most of the victims were black or brown, and the youth population between the ages of 18 and 30 is the most affected by violent crime. Lastly, it should be noted that about 88% of the criminal events under analysis involved firearms. Figure \ref{fig:plot-crime} shows the spatial distribution of the 1,657 LIVC crimes that occurred in the city of Recife between July, 2008 and June, 2010.

@@

sed_descriptives <- data.frame(variable = c("Median Income", 
                                            "Proportion Unemployed", 
                                            "Proportion Poverty", 
                                            "Proportion Single Mother", 
                                            "Proportion Young Single Mother",
                                            "Population Density"),
                               min = c(min(setores.sf$median_income),
                                       min(setores.sf$prop_unemployed),
                                       min(setores.sf$prop_poverty),
                                       min(setores.sf$prop_single_mother),
                                       min(setores.sf$prop_young_single_mother),
                                       min(setores.sf$pop_density)),
                               second_q = c(quantile(setores.sf$median_income, c(0, 0.25, 0.5, 0.75, 1))[2],
                                            quantile(setores.sf$prop_unemployed, c(0, 0.25, 0.5, 0.75, 1))[2],
                                            quantile(setores.sf$prop_poverty, c(0, 0.25, 0.5, 0.75, 1))[2],
                                            quantile(setores.sf$prop_single_mother, c(0, 0.25, 0.5, 0.75, 1))[2],
                                            quantile(setores.sf$prop_young_single_mother, c(0, 0.25, 0.5, 0.75, 1))[2],
                                            quantile(setores.sf$pop_density, c(0, 0.25, 0.5, 0.75, 1))[2]),
                               median = c(median(setores.sf$median_income),
                                          median(setores.sf$prop_unemployed),
                                          median(setores.sf$prop_poverty),
                                          median(setores.sf$prop_single_mother),
                                          median(setores.sf$prop_young_single_mother),
                                          median(setores.sf$pop_density)),
                               mean = c(mean(setores.sf$median_income),
                                        mean(setores.sf$prop_unemployed),
                                        mean(setores.sf$prop_poverty),
                                        mean(setores.sf$prop_single_mother),
                                        mean(setores.sf$prop_young_single_mother),
                                        mean(setores.sf$pop_density)),
                               third_q = c(quantile(setores.sf$median_income, c(0, 0.25, 0.5, 0.75, 1))[4],
                                           quantile(setores.sf$prop_unemployed, c(0, 0.25, 0.5, 0.75, 1))[4],
                                           quantile(setores.sf$prop_poverty, c(0, 0.25, 0.5, 0.75, 1))[4],
                                           quantile(setores.sf$prop_single_mother, c(0, 0.25, 0.5, 0.75, 1))[4],
                                           quantile(setores.sf$prop_young_single_mother, c(0, 0.25, 0.5, 0.75, 1))[4],
                                           quantile(setores.sf$pop_density, c(0, 0.25, 0.5, 0.75, 1))[4]),
                               max = c(max(setores.sf$median_income),
                                           max(setores.sf$prop_unemployed),
                                           max(setores.sf$prop_poverty),
                                           max(setores.sf$prop_single_mother),
                                           max(setores.sf$prop_young_single_mother),
                                           max(setores.sf$pop_density)),
                               pca_loadings = as.numeric(setores.pca$rotation[c(2, 3, 6, 4, 5, 1), 1]))
```

@@

As seen in the table, Recife is a city with large socio-economic and demographic disparities; for example, the _setor_ with the highest median income has a median income that is `r prettyNum(filter(sed_descriptives, variable == "Median Income")["max"]/filter(sed_descriptives, variable == "Median Income")["min"] * 100, digits = 2, big.mark = ",")`% higher than the median income in the _setor_ with the lowest median income. Whereas there are _setores_ where the proportion of population who are unemployed is zero, there are _setores_ where almost 60% of the population are unemployed (which for some may mean they are part of the informal econom). Likewise, there are _setores_ where almost 70% of the population lives in poverty. In addition to these economic indicators, two variables are used to represent family disruption, the proportion of families whose head is a single mother, and the proportion of families whose head is a _young_ single mother, that is, a woman between the ages of 15 and 25. As can be seen, there are _setores_ with approximately 70% of households led by single mothers, and of these, over 21% are led by younger women, indicating a high degree of family disruption. Unlike other places, where higher population density is associated with more poverty, the distribution of this variable in Recife is more mixed: middle income and higher income households favor high density development in many parts of the city. The spatial distribution of these variables can be seen in Figure \ref{fig:plot-sed-variables}.

@@

Once the indicator of Socio-Economic Deprivation has been obtained, as outlined above, it can be used for conditional quadrat analysis. As previously described, the quadrats are irregular areas that are defined based on a covariate, in this case the quintiles of the Socio-Economic Deprivation indicator. Each color in Figure \ref{fig:plot-sed-as-quintiles} is a "container" for counting events. In essence, this entails counting the number of events at each level of the covariate, and then calculating an estimator of the intensity. The global intensity of each type of event, as well as the quadrant counts and intensity at each level of SED can be consulted in Table \ref{tab:quadrat-count-results}.

@@

intensity_all <- data.frame(SED = rep(c("Low", 
                                        "Middle Low", 
                                        "Middle", 
                                        "Middle High", 
                                        "High"), 
                                      times = 7),
                            Type = rep(c("Violent Crime", 
                                         "Catholic",
                                         "Evangelical",
                                         "Spiritist",
                                         "Ice Cream",
                                         "Pharmacies",
                                         "Super Markets"), 
                                       each = 5),
                            Quadrat_Count = c(crime5,
                                              catholic5,
                                              evangelic5,
                                              spiritist5,
                                              ice_cream5,
                                              pharmacies5,
                                              super_markets5),
                            Intensity = c(crime5_intensity,
                                          catholic5_intensity,
                                          evangelic5_intensity,
                                          spiritist5_intensity,
                                          ice_cream5_intensity,
                                          pharmacies5_intensity,
                                          supermarkets5_intensity),
                            Global_Intensity = rep(c(summary_crime$intensity,
                                                     summary_catholic$intensity,
                                                     summary_evangelic$intensity,
                                                     summary_spiritist$intensity + 1e-8, #small correction to distinguish from ice cream in table
                                                     summary_ice_cream$intensity,
                                                     summary_pharmacy$intensity,
                                                     summary_supermarket$intensity),
                                                   each = 5)) %>%
  mutate(SED = factor(SED, levels = c("Low", 
                                      "Middle Low", 
                                      "Middle", 
                                      "Middle High", 
                                      "High"),
                      ordered = TRUE),
         Type = factor(Type, levels = c("Violent Crime", 
                                        "Catholic",
                                        "Evangelical",
                                        "Spiritist",
                                        "Ice Cream",
                                        "Pharmacies",
                                        "Super Markets")))
```

```{r table-counts-intensity-summary, echo=FALSE}
kable(intensity_all %>% 
        dplyr::select(Type, Global_Intensity, SED, Quadrat_Count, Intensity) %>% 
        mutate(Global_Intensity = prettyNum(Global_Intensity, digits = 3),
               Intensity = prettyNum(Intensity, digits = 3)),
      "latex",
      booktabs = TRUE,
      #digits = 3,
      escape = FALSE,
      align = c("l", "c", "c", "c"),
      caption = "\\label{tab:quadrat-count-results}Conditional quadrat counts results",
      col.names = linebreak(c("Event", 
                              "Global Intensity", 
                              "SED Quintile", 
                              "Quadrat Count",
                              "Intensity"))) %>%
  kable_styling(bootstrap_options = c("striped", "scaled_down")) %>% 
  collapse_rows(1:2, latex_hline = "major")
```

### 2020-06-13 Moral-Communities-and-Crime/Moral_Communities_and_Crime.tex
v2 Second submission after comments from Edivaldo and Tatiane

The case study is the city of Recife in the state of Pernambuco, in
Brazil's Northeast. Recife is a large and important metropolitan area in
a historically poor region, and afflicted by high levels of violent
crime, having the dubious honor of being one of five state capitals with
the highest rates of homicide in the period under study (Menezes et al.
2013). The empirical strategy is to use spatial analysis to explore the
potential geographical relationships between violent criminal events, on
the one hand, and places of worship and a selection of commercial
establishments that serve as controls, on the other.

@@

The data are organized at the individual level and it is possible to
obtain information about the crime event location, day of the week, day
of the month, period of the day, as well as gender, age, and race of the
victim. The database used in this study comprises the period from July
1, 2008 to June 30, 2010. Some descriptive statistics regarding this
dataset are reported in Table 1, where it can be seen that overall,
about 91\% of the victims of LIVC in the period analyzed were men, while
approximately 98\% of violent crimes were homicides. In addition, most
of the victims were black or brown, and the youth population between the
ages of 18 and 30 is the most affected by violent crime. Lastly, it
should be noted that about 88\% of the criminal events under analysis
involved firearms. Figure \ref{fig:plot-crime} shows the spatial
distribution of the 1,657 LIVC crimes that occurred in the city of
Recife between July, 2008 and June, 2010.

@@

\caption{\label{tab:table-sed-descriptive-statistics}\label{tab:sed-descriptive-statistics}Descriptive statistics of some key socio-economic and demographic variables in Recife}
\centering
\begin{tabular}[t]{lccccccc}
\toprule
Variable & Min & 2nd Quartile & Median & Mean & 3rd Quartile & Max & \makecell[l]{PC Factor 1\\ Loadings}\\
\midrule
Median Income & 122.05 & 337.40 & 581.67 & 1031.09 & 1283.79 & 181642.5 & -0.42\\
Proportion Unemployed & 0.00 & 0.05 & 0.10 & 0.10 & 0.14 & 181642.5 & 0.56\\
Proportion Poverty & 0.00 & 0.05 & 0.11 & 0.12 & 0.17 & 181642.5 & 0.58\\
Proportion Single Mother & 0.00 & 0.33 & 0.37 & 0.37 & 0.41 & 181642.5 & 0.17\\
Proportion Young Single Mother & 0.00 & 0.01 & 0.02 & 0.02 & 0.02 & 181642.5 & 0.38\\
\addlinespace
Population Density & 0.00 & 10564.16 & 16333.62 & 19230.23 & 25106.07 & 181642.5 & 0.03\\
\bottomrule
\multicolumn{8}{l}{\textit{Note: }}\\
\multicolumn{8}{l}{The variable for young single mothers is for women aged 15-25}\\
\multicolumn{8}{l}{The first principal component accounts for 44.9\% of the variance}\\
\end{tabular}
\end{table}

As seen in the table, Recife is a city with large socio-economic and
demographic disparities; for example, the \emph{setor} with the highest
median income has a median income that is 148,826\% higher than the
median income in the \emph{setor} with the lowest median income. Whereas
there are \emph{setores} where the proportion of population who are
unemployed is zero, there are \emph{setores} where almost 60\% of the
population are unemployed. Likewise, there are \emph{setores} where
almost 70\% of the population lives in poverty. In addition to these
economic indicators, two variables are used to represent family
disruption, the proportion of families whose head is a single mother,
and the proportion of families whose head is a \emph{young} single
mother, that is, a woman between the ages of 15 and 25. As can be seen,
there are \emph{setores} with approximately 70\% of households led by
single mothers, and of these, over 21\% are led by younger women,
indicating a high degree of family disruption. Unlike other places,
where higher population density is associated with more poverty, the
distribution of this variable in Recife is more mixed: middle income and
higher income households favor high density development in many parts of
the city. The spatial distribution of these variables can be seen in
Figure \ref{fig:plot-sed-variables}.

@@

Once the indicator of Socio-Economic Deprivation has been obtained, as
outlined above, it can be used for conditional quadrat analysis. As
previously described, the quadrats are irregular areas that are defined
based on a covariate, in this case the quintiles of the Socio-Economic
Deprivation indicator. Each color in Figure
\ref{fig:plot-sed-as-quintiles} is a ``container'' for counting events.
In essence, this entails counting the number of events at each level of
the covariate, and then calculating an estimator of the intensity. The
global intensity of each type of event (i.e., places of worship and
businesses), as well as quadrant counts and intensity at each level of
SED can be consulted in Table \ref{tab:quadrat-count-results}.

@@

\caption{\label{tab:table-counts-intensity-summary}\label{tab:quadrat-count-results}Conditional quadrat counts results}
\centering
\begin{tabular}[t]{lcccl}
\toprule
Event & Global Intensity & SED Quintile & Quadrat Count & Intensity\\
\midrule
 &  & Low & 201 & 0\\

 &  & Middle Low & 295 & 0\\

 &  & Middle & 364 & 0\\

 &  & Middle High & 479 & 0\\

\multirow{-5}{*}{\raggedright\arraybackslash Violent Crime} &  & High & 318 & 0\\

 &  & Low & 32 & 0\\

 &  & Middle Low & 26 & 0\\

 &  & Middle & 33 & 0\\

 &  & Middle High & 29 & 0\\

\multirow{-5}{*}{\raggedright\arraybackslash Catholic} &  & High & 11 & 0\\

 &  & Low & 209 & 0\\

 &  & Middle Low & 304 & 0\\

 &  & Middle & 334 & 0\\

 &  & Middle High & 345 & 0\\

\multirow{-5}{*}{\raggedright\arraybackslash Evangelical} &  & High & 185 & 0\\

 &  & Low & 37 & 0\\

 &  & Middle Low & 50 & 0\\

 &  & Middle & 40 & 0\\

 &  & Middle High & 40 & 0\\

\multirow{-5}{*}{\raggedright\arraybackslash Spiritist} &  & High & 20 & 0\\

 &  & Low & 45 & 0\\

 &  & Middle Low & 60 & 0\\

 &  & Middle & 39 & 0\\

 &  & Middle High & 37 & 0\\

\multirow{-5}{*}{\raggedright\arraybackslash Ice Cream} &  & High & 26 & 0\\

 &  & Low & 116 & 0\\

 &  & Middle Low & 122 & 0\\

 &  & Middle & 91 & 0\\

 &  & Middle High & 84 & 0\\

\multirow{-5}{*}{\raggedright\arraybackslash Pharmacies} &  & High & 25 & 0\\

 &  & Low & 27 & 0\\

 &  & Middle Low & 32 & 0\\

 &  & Middle & 38 & 0\\

 &  & Middle High & 16 & 0\\

\multirow{-5}{*}{\raggedright\arraybackslash Super Markets} & \multirow{-35}{*}{\centering\arraybackslash 0} & High & 12 & 0\\
\bottomrule
\end{tabular}
\end{table}

>>>>

The empirical analysis uses geographically detailed data from the city
of Recife in the state of Pernambuco, in Brazil's Northeast. Recife is a
large and important metropolitan area in a historically poor region, and
afflicted by high levels of violent crime, having the dubious honor of
being one of five state capitals with the highest rates of homicide in
the period under study (Menezes et al. 2013). The empirical strategy is
to use spatial analysis to explore the potential geographical
relationships between violent criminal events, on the one hand, and
places of worship and a selection of commercial establishments that
serve as controls, on the other.

@@

The data are organized at the individual level and it is possible to
obtain information about the crime event location, day of the week, day
of the month, period of the day, as well as gender, age, and race of the
victim. The database used in this study comprises the period from July
1, 2008 to June 30, 2010. Some descriptive statistics regarding this
dataset are reported in Table 1, where it can be seen that overall,
about 91\% of the victims of LIVC in the period analyzed were men. In
addition, most of the victims were black or brown, and the youth
population between the ages of 18 and 30 is the most affected by violent
crime. Lastly, it should be noted that about 88\% of the criminal events
under analysis involved firearms. Figure \ref{fig:plot-crime} shows the
spatial distribution of the 1,657 LIVC crimes that occurred in the city
of Recife between July, 2008 and June, 2010.

@@

\caption{\label{tab:table-sed-descriptive-statistics}\label{tab:sed-descriptive-statistics}Descriptive statistics of some key socio-economic and demographic variables in Recife}
\centering
\begin{tabular}[t]{lccccccc}
\toprule
Variable & Min & 2nd Quartile & Median & Mean & 3rd Quartile & Max & \makecell[l]{PC Factor 1\\ Loadings}\\
\midrule
Median Income & 122.05 & 337.40 & 581.67 & 1031.09 & 1283.79 & 7954.19 & -0.42\\
Proportion Unemployed & 0.00 & 0.05 & 0.10 & 0.10 & 0.14 & 0.59 & 0.56\\
Proportion Poverty & 0.00 & 0.05 & 0.11 & 0.12 & 0.17 & 0.63 & 0.58\\
Proportion Single Mother & 0.00 & 0.33 & 0.37 & 0.37 & 0.41 & 0.70 & 0.17\\
Proportion Young Single Mother & 0.00 & 0.01 & 0.02 & 0.02 & 0.02 & 0.21 & 0.38\\
\addlinespace
Population Density & 0.00 & 10564.16 & 16333.62 & 19230.23 & 25106.07 & 181642.55 & 0.03\\
\bottomrule
\multicolumn{8}{l}{\textit{Note: }}\\
\multicolumn{8}{l}{The variable for young single mothers is for women aged 15-25}\\
\multicolumn{8}{l}{The first principal component accounts for 44.9\% of the variance}\\
\end{tabular}
\end{table}

As seen in the table, Recife is a city with large socio-economic and
demographic disparities; for example, the \emph{setor} with the highest
median income has a median income that is 6,517\% higher than the median
income in the \emph{setor} with the lowest median income. Whereas there
are \emph{setores} where the proportion of population who are unemployed
is zero, there are \emph{setores} where almost 60\% of the population
are unemployed (which for some may mean they are part of the informal
econom). Likewise, there are \emph{setores} where almost 70\% of the
population lives in poverty. In addition to these economic indicators,
two variables are used to represent family disruption, the proportion of
families whose head is a single mother, and the proportion of families
whose head is a \emph{young} single mother, that is, a woman between the
ages of 15 and 25. As can be seen, there are \emph{setores} with
approximately 70\% of households led by single mothers, and of these,
over 21\% are led by younger women, indicating a high degree of family
disruption. Unlike other places, where higher population density is
associated with more poverty, the distribution of this variable in
Recife is more mixed: middle income and higher income households favor
high density development in many parts of the city. The spatial
distribution of these variables can be seen in Figure
\ref{fig:plot-sed-variables}.

@@

Once the indicator of Socio-Economic Deprivation has been obtained, as
outlined above, it can be used for conditional quadrat analysis. As
previously described, the quadrats are irregular areas that are defined
based on a covariate, in this case the quintiles of the Socio-Economic
Deprivation indicator. Each color in Figure
\ref{fig:plot-sed-as-quintiles} is a ``container'' for counting events.
In essence, this entails counting the number of events at each level of
the covariate, and then calculating an estimator of the intensity. The
global intensity of each type of event, as well as the quadrant counts
and intensity at each level of SED can be consulted in Table
\ref{tab:quadrat-count-results}.

@@

\caption{\label{tab:table-counts-intensity-summary}\label{tab:quadrat-count-results}Conditional quadrat counts results}
\centering
\begin{tabular}[t]{lcccl}
\toprule
Event & Global Intensity & SED Quintile & Quadrat Count & Intensity\\
\midrule
 &  & Low & 201 & 4.6e-06\\

 &  & Middle Low & 295 & 6.75e-06\\

 &  & Middle & 364 & 8.33e-06\\

 &  & Middle High & 479 & 8.65e-06\\

\multirow{-5}{*}{\raggedright\arraybackslash Violent Crime} & \multirow{-5}{*}{\centering\arraybackslash 7.58e-06} & High & 318 & 9.94e-06\\
\cmidrule{1-5}
 &  & Low & 32 & 7.32e-07\\

 &  & Middle Low & 26 & 5.95e-07\\

 &  & Middle & 33 & 7.55e-07\\

 &  & Middle High & 29 & 5.24e-07\\

\multirow{-5}{*}{\raggedright\arraybackslash Catholic} & \multirow{-5}{*}{\centering\arraybackslash 5.99e-07} & High & 11 & 3.44e-07\\
\cmidrule{1-5}
 &  & Low & 209 & 4.78e-06\\

 &  & Middle Low & 304 & 6.95e-06\\

 &  & Middle & 334 & 7.65e-06\\

 &  & Middle High & 345 & 6.23e-06\\

\multirow{-5}{*}{\raggedright\arraybackslash Evangelical} & \multirow{-5}{*}{\centering\arraybackslash 6.3e-06} & High & 185 & 5.78e-06\\
\cmidrule{1-5}
 &  & Low & 37 & 8.46e-07\\

 &  & Middle Low & 50 & 1.14e-06\\

 &  & Middle & 40 & 9.16e-07\\

 &  & Middle High & 40 & 7.22e-07\\

\multirow{-5}{*}{\raggedright\arraybackslash Spiritist} & \multirow{-5}{*}{\centering\arraybackslash 8.65e-07} & High & 20 & 6.25e-07\\
\cmidrule{1-5}
 &  & Low & 45 & 1.03e-06\\

 &  & Middle Low & 60 & 1.37e-06\\

 &  & Middle & 39 & 8.93e-07\\

 &  & Middle High & 37 & 6.68e-07\\

\multirow{-5}{*}{\raggedright\arraybackslash Ice Cream} & \multirow{-5}{*}{\centering\arraybackslash 9.47e-07} & High & 26 & 8.12e-07\\
\cmidrule{1-5}
 &  & Low & 116 & 2.65e-06\\

 &  & Middle Low & 122 & 2.79e-06\\

 &  & Middle & 91 & 2.08e-06\\

 &  & Middle High & 84 & 1.52e-06\\

\multirow{-5}{*}{\raggedright\arraybackslash Pharmacies} & \multirow{-5}{*}{\centering\arraybackslash 2e-06} & High & 25 & 7.81e-07\\
\cmidrule{1-5}
 &  & Low & 27 & 6.17e-07\\

 &  & Middle Low & 32 & 7.32e-07\\

 &  & Middle & 38 & 8.7e-07\\

 &  & Middle High & 16 & 2.89e-07\\

\multirow{-5}{*}{\raggedright\arraybackslash Super Markets} & \multirow{-5}{*}{\centering\arraybackslash 5.72e-07} & High & 12 & 3.75e-07\\
\bottomrule
\end{tabular}
\end{table}
